% SynoptiQ: Project Overview
% Compile with: xelatex project_overview.tex
\documentclass[11pt,letterpaper]{article}

% ── Core packages ───────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage{fontspec}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage[table]{xcolor}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{csquotes}
\usepackage[letterpaper, top=0.9in, bottom=1.0in, left=1.0in, right=1.0in]{geometry}
\usepackage{titlesec}
\usepackage{fancyhdr}
\usepackage{tcolorbox}
\tcbuselibrary{skins,breakable}
\usepackage{array}
\usepackage{tabularx}
\usepackage{ragged2e}
\usepackage{float}
\usepackage[font={small,color=stone},labelfont={bf,color=marine},labelsep=quad]{caption}

\setlist{nosep, leftmargin=*}
\frenchspacing
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.55em}

% ── Typography ─────────────────────────────────────────────────────
\setmainfont{TeX Gyre Pagella}[Ligatures=TeX]
\newfontfamily\headfont{Poppins}
\newfontfamily\greekfont{FreeSerif}
\newcommand{\gk}[1]{{\greekfont\itshape #1}}
\setmonofont{Latin Modern Mono}[Scale=0.92]

% ── Palette ────────────────────────────────────────────────────────
\definecolor{marine}{HTML}{123C48}
\definecolor{marinelight}{HTML}{2E6272}
\definecolor{terracotta}{HTML}{B75B36}
\definecolor{papyrus}{HTML}{F7F2E7}
\definecolor{stone}{HTML}{6E6A5F}
\definecolor{rulegrey}{HTML}{D9D2C0}
\definecolor{inkblack}{HTML}{201D1A}
\color{inkblack}
\arrayrulecolor{marine!65!black}

\newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}}
\newcolumntype{L}[1]{>{\RaggedRight\arraybackslash}p{#1}}
\renewcommand\labelitemi{\textcolor{terracotta}{\textbullet}}

\hypersetup{
  colorlinks=true,
  linkcolor=marine,
  citecolor=marine,
  urlcolor=terracotta,
  pdftitle={SynoptiQ: A Computational Framework for the Synoptic Problem},
  pdfauthor={Abderahmane Ainouche}
}

% ── Section styling ─────────────────────────────────────────────────
\titleformat{\section}
  {\headfont\Large\bfseries\color{marine}}
  {\color{marine}\thesection}
  {0.65em}{}
  [{\vspace{2pt}\color{rulegrey}\titlerule[0.9pt]}]
\titlespacing*{\section}{0pt}{1.7em}{0.9em}

\titleformat{\subsection}
  {\headfont\large\bfseries\color{marinelight}}
  {\thesubsection}{0.6em}{}
\titlespacing*{\subsection}{0pt}{1.3em}{0.5em}

% ── Header / footer ─────────────────────────────────────────────────
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\footnotesize\headfont\color{stone} SynoptiQ — Project Overview}
\fancyhead[R]{\footnotesize\headfont\color{stone} Abderahmane Ainouche}
\fancyfoot[C]{\footnotesize\headfont\color{stone}\thepage}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\headrule}{{\color{rulegrey}\hrule height\headrulewidth}}
\fancypagestyle{firstpage}{%
  \fancyhf{}
  \fancyfoot[C]{\footnotesize\headfont\color{stone}\thepage}
  \renewcommand{\headrulewidth}{0pt}
}

% ── Helpers ─────────────────────────────────────────────────────────
\newcommand{\statcard}[2]{%
  \cellcolor{papyrus}{\begin{minipage}[c][2.05cm][c]{\linewidth}\centering
    {\headfont\bfseries\fontsize{17}{19}\selectfont\color{marine}#1}\\[3pt]
    {\footnotesize\headfont\color{stone}#2}
  \end{minipage}}%
}
\newcommand{\tblhead}{\rowcolor{marine!12}}

\begin{document}

% ── Title block ──────────────────────────────────────────────────────
\thispagestyle{firstpage}
\begin{flushleft}
{\headfont\footnotesize\bfseries\color{terracotta}\MakeUppercase{Project Overview}}\\[7pt]
{\headfont\bfseries\fontsize{22}{27}\selectfont\color{marine}%
SynoptiQ: A Computational Framework\\[4pt] for the Synoptic Problem}\\[10pt]
{\headfont\large\color{stone} Abderahmane Ainouche}
\end{flushleft}
\vspace{6pt}
{\color{rulegrey}\rule{\linewidth}{0.9pt}}
\vspace{1.3em}

% ── Summary box ─────────────────────────────────────────────────────
\begin{tcolorbox}[
  enhanced, breakable,
  colback=papyrus, colframe=papyrus,
  boxrule=0pt, arc=0pt, outer arc=0pt,
  borderline west={2.6pt}{0pt}{marine},
  left=16pt, right=16pt, top=11pt, bottom=11pt
]
{\headfont\bfseries\color{marine} In Brief}\par\vspace{5pt}
\noindent For over two centuries, scholars have debated how the Gospels of
Matthew, Mark, and Luke are literarily related---who copied from whom, and
what lost sources might lie behind the text.  \textbf{SynoptiQ} is a
multi-phase computational framework that brings modern neural methods to
bear on this question: transformer-based language models trained on Koine
Greek, neural direction-of-copying detection, editorial fatigue modeling,
Q reconstruction, and Bayesian model comparison across the major Synoptic
hypotheses.  The first phase---a domain-adapted Koine Greek language model
and a morphologically annotated corpus of the Synoptic Gospels---is
complete and released under open licenses.  Two further phases are in
active development.
\end{tcolorbox}

\vspace{1.1em}
\begin{center}
\renewcommand{\arraystretch}{1}
\setlength{\tabcolsep}{7pt}
\begin{tabular}{P{0.18\linewidth}P{0.18\linewidth}P{0.18\linewidth}P{0.18\linewidth}P{0.18\linewidth}}
\statcard{3}{Papers}
& \statcard{7}{Phases}
& \statcard{49{,}061}{Annotated tokens}
& \statcard{170}{Pericopes}
& \statcard{Open}{Source}
\end{tabular}
\end{center}
\vspace{0.6em}

\section{The Question}

The Synoptic Gospels---Matthew, Mark, and Luke---share extensive verbal
parallels.  Over 90\% of Mark's verses have a parallel in Matthew, and
over half appear in Luke.  The dominant explanation, the Two-Source
Hypothesis, holds that Matthew and Luke independently used Mark and a
lost sayings source called Q.  But alternatives---the Farrer--Goulder
hypothesis (Luke used Matthew directly, no Q), the Augustinian hypothesis
(Matthew wrote first, Mark abbreviated), and the Griesbach hypothesis
(two-gospel)---each have scholarly defenders.

The evidence is textual, multi-layered, and resists simple statistical
treatment.  A scribe copying a source may follow it closely in one verse
and diverge in the next.  Editorial fatigue---the tendency of a copying
author to revert to their own stylistic norms over the course of a
pericope---leaves subtle but detectable traces.  And the direction of
literary dependence between any two parallel passages is often unclear
from surface features alone.

Computational methods offer a new way forward.  But applying them
requires building the infrastructure first: a machine-readable corpus,
a language model that understands the linguistic register of the texts,
and a suite of neural tools designed specifically for textual criticism.

\textbf{SynoptiQ is that infrastructure.}

\section{The Framework}

SynoptiQ proceeds in three stages, each yielding a standalone paper.
The full pipeline is: \textit{corpus and language model $\rightarrow$
direction detection and editorial fatigue $\rightarrow$ source
reconstruction and Bayesian comparison.}

\subsection{Paper A --- Foundation: Corpus \& Language Model \hfill \textit{Complete}}

The Synoptiq Corpus is a token-aligned, morphologically annotated dataset
of the Synoptic Gospels: 49,061 tokens across 170 Aland pericopes, with
235 pairwise Needleman--Wunsch token alignments and 13-class CCAT
part-of-speech annotations.  It is released as a HuggingFace dataset with
stratified train/validation/test splits.

KoineFormer is a domain-adapted T5 encoder-decoder trained on 1.5M tokens
of Koine Greek (SBLGNT + Apostolic Fathers) with a Classical Greek replay
buffer.  Using LoRA adapters---only 3.7M of the 220M parameters are
trained---KoineFormer substantially outperforms the zero-shot Classical
Greek baseline on part-of-speech tagging while producing a compact 14 MB
checkpoint.  The accompanying paper provides full evaluation and analysis.

\medskip
\begin{tcolorbox}[
  enhanced, breakable,
  colback=marine!6, colframe=marine!30,
  boxrule=0.5pt, arc=3pt,
  left=12pt, right=12pt, top=8pt, bottom=8pt
]
{\headfont\bfseries\color{marine} Released}\\[4pt]
\begin{tabular}{@{}L{4cm}L{5.5cm}L{3cm}@{}}
\toprule
\tblhead\textbf{Artifact} & \textbf{Location} & \textbf{License} \\
\midrule
KoineFormer model & hf.co/ainouche-abderahmane/koineformer & CC-BY-SA 4.0 \\
Synoptiq Corpus & hf.co/datasets/ainouche-abderahmane/synoptiq-corpus & CC-BY-SA 4.0 \\
Full codebase & github.com/abderahmane-ai/SynoptiQ & CC-BY-SA 4.0 \\
\bottomrule
\end{tabular}
\end{tcolorbox}

\subsection{Paper B --- Direction Detection \& Editorial Fatigue \hfill \textit{In Development}}

The core question---``who copied from whom?''---is addressed with two
complementary neural methods.

\textbf{Direction Scorer.}  Frozen KoineFormer token states produce ten
calibrated asymmetry features fed into a swap-equivariant 3-way classifier
(A$\rightarrow$B, B$\rightarrow$A, independent).  The classifier emits
logits $[d, -d, i]$, so reversing the passage order swaps only the two
directional labels while preserving the independent label.
Training uses triple-tradition pericopes where the direction is known
(Mark$\rightarrow$Matthew, Mark$\rightarrow$Luke under the consensus
view).

\textbf{Editorial Fatigue Model.}  A position-weighted consistency measure
quantifies the well-documented phenomenon that a copying author tends to
revert to their own stylistic norms over the course of a pericope---the
model assigns greater weight to early verses, where fidelity to the
source is highest, and tracks the divergence as the author's own habits
reassert themselves.  This provides converging evidence from a signal
independent of the direction scorer.

\subsection{Paper C --- Q Reconstruction \& Bayesian Comparison \hfill \textit{Planned}}

\textbf{Q Reconstruction.}  A Fusion-in-Decoder architecture encodes
Matthew and Luke independently, concatenates their hidden states, and
uses a decoder with cross-attention to reconstruct the shared source.
Training proceeds in two stages: first on triple tradition (Matthew +
Luke $\rightarrow$ reconstruct Mark, where we have ground truth), then
transferred to double tradition for the hypothesized Q material.

\textbf{Bayesian Model Comparison.}  Direction scorer outputs---with
per-pericope uncertainty estimates obtained via Monte Carlo
dropout---feed hierarchical Bayesian models representing the four major
hypotheses: Two-Source (2SH), Farrer--Goulder (FGH), Augustinian, and
Griesbach.  Bridge sampling yields Bayes factors for formal model
comparison, treating the Synoptic Problem as a probabilistic question
rather than a binary debate.

\section{What Distinguishes This Work}

Most computational approaches to the Synoptic Problem have used
surface-level features: word frequencies, n-gram overlap, stylometric
distances.  SynoptiQ operates at a deeper level:

\begin{itemize}
\item \textbf{Contextual representations.}  Transformer hidden states
  capture syntactic and semantic patterns that surface features miss.
\item \textbf{Causal direction detection.}  Frozen-encoder asymmetry
  features provide a directional signal, rather than symmetric similarity
  measures.
\item \textbf{Swap-equivariant classification.}  The direction head
  enforces the known label symmetry: swapping A and B must swap
  A$\rightarrow$B and B$\rightarrow$A while leaving independence unchanged.
\item \textbf{Full uncertainty quantification.}  Bayesian hierarchical
  models with bridge sampling yield Bayes factors and credible intervals,
  not just point estimates.
\item \textbf{Complete transparency.}  Model weights, dataset, training
  code, evaluation scripts, and loss curves are all public.
\end{itemize}

\section{Current Status and Timeline}

\begin{table}[H]
\centering
\begin{tabular}{lL{3.5cm}L{5cm}l}
\toprule
\tblhead\textbf{Phase} & \textbf{Component} & \textbf{Key Deliverable} & \textbf{Status} \\
\midrule
1 & Corpus \& alignment & Synoptiq Corpus (HF dataset) & \checkmark\ Complete \\
2A & KoineFormer DAPT & Domain-adapted LM (HF model) & \checkmark\ Complete \\
2B & Multi-task fine-tuning & POS, parser, pericope heads & Code ready \\
3 & Direction scorer & 10-feature swap-equivariant classifier & Code ready \\
4 & Editorial fatigue & Position-weighted KL model & Designed \\
5 & Q reconstruction & FiD transfer learning & Architecture ready \\
6 & Bayesian comparison & PyMC hierarchical + BF & Specified \\
7 & Interpretability & SHAP + BERTViz + Hawkins & Specified \\
\bottomrule
\end{tabular}
\caption{Phases of the SynoptiQ project.  Phase~2A is complete and documented in the accompanying paper (\texttt{main.tex}).  Phases~2B--7 are in various stages of development.}
\end{table}

\section{Why This Matters}

The Synoptic Problem is not merely an antiquarian puzzle.  How one
resolves it affects how one reads the Gospels---which passages are
independent witnesses, which are literary reworkings, and what can be
inferred about the historical Jesus.  A computational framework that
brings rigorous, reproducible evidence to bear on these questions serves
both New Testament scholarship and the broader digital humanities.

SynoptiQ is designed to be \textit{cumulative}.  Each phase produces
standalone results, but the full value emerges from the integration:
directional evidence from Paper~B feeds the Bayesian priors of Paper~C;
the encoder from Paper~A underlies everything.  The project is open from
the start---model weights, dataset, and code are already public---so that
scholars can inspect, reproduce, and build on each stage.

\vspace{1em}
\begin{flushright}
\textit{--- Abderahmane Ainouche}\\[4pt]
{\footnotesize\headfont\color{stone}
github.com/abderahmane-ai/SynoptiQ \hspace{1.5em}
ainouche-abderahmane/synoptiq-corpus (HF)}
\end{flushright}

\end{document}
